Gravitational-wave sensitivity curves

### 2014-07-08 Detectors.tex
Odds and Ends

Ground-based detectors are the most numerous. A collection of interferometric detectors are listed in table \ref{table:t}, these are sensitive to GWs in the frequency range ${\mathcal{O}}(10$--$10^{3})~\mathrm{Hz}$. They all simulate free-floating test masses by suspending a mass from a pendulum system with natural frequency much greater than that of the GW. Their sensitivity curves include narrow lines, which arise from noise sources in the instrument including resonances in the suspension system and electrical noise at multiples of $60~\mathrm{Hz}$: these have been removed in the \ref{app:a} figures for clarity. The detectors fall broadly into three categories: first-generation detectors, which have already operated; second-generation detectors currently under construction, and third-generation detectors at the planning stage.

The most notable ground-based detectors are LIGO and Virgo \citep{Abramovici1992,Acernese2008a}. These work in collaboration, supported by GEO600 \citep{Willke2002}. LIGO and Virgo have completed science runs as first-generation detectors, now referred to as Initial LIGO (iLIGO) and Initial Virgo (iVirgo). They are now being upgraded to second-generation detectors, advanced LIGO (aLIGO) and advanced Virgo (AdV). LIGO has two observatories, one at Hanford, Washington, which has two detectors, and another at Livingston, Louisiana. There is an agreement to move one of the upgraded Hanford detector systems to a location in India \citep{LIGO-India,Unnikrishnan2013}. The advanced detectors should start operation in the next couple of years, with LIGO-India following further in the future.

TAMA300 is a Japanese first-generation detector. Its successor is the Kamioka Gravitational Wave Detector (KAGRA), formerly the Large-scale Cryogenic Gravitational wave Telescope (LCGT), which is located underground in the Kamioka mine. It employs more sophisticated noise-reduction techniques than LIGO or Virgo, such as cryogenic cooling \citep{2012CQGra..29l4007S}.

The Einstein Telescope (ET) is an ambitious proposal to construct an underground third-generation detector \citep{Punturo2010,Sathyaprakash2012}. Its location would provide shielding from seismic noise, allowing it to observe frequencies of $10$--$10^4~\mathrm{Hz}$.

@@

We use an interpolation to the data published on \url{https://wwwcascina.virgo.infn.it/advirgo/} (2013) for the AdV sensitivity curve, an interpolation to the data for version D of the detector published on \url{http://gwcenter.icrr.u-tokyo.ac.jp/en/researcher/parameter} (2013) for KAGRA and analytic fits to the sensitivity curves from \citet{Sathyaprakash} for the remaining detectors.

@@

These missions are designed to probe the decihertz region of the GW spectrum; they are considerably more ambitious than the LISA or eLISA mission and their launches will be further into the future. We use a simple analytic fit to the sensitivity curve for ALIA \cplb{[citation needed]}, while for DECIGO and BBO, fits to the sensitivity curves given by \citet{2011PhRvD..83d4011Y} are used.

>>>>

Ground-based detectors are the most numerous. A collection of interferometric detectors are listed in table \ref{table:t}, these are sensitive to GWs in the frequency range ${\mathcal{O}}(10$--$10^{3})~\mathrm{Hz}$. They all simulate free-floating test masses by suspending a mass from a pendulum system with natural frequency much greater than that of the GW. Their sensitivity curves include narrow lines that arise from noise sources in the instrument, including resonances in the suspension system and electrical noise at multiples of $60~\mathrm{Hz}$: these have been removed in the \ref{app:a} figures for clarity. The detectors fall broadly into three categories: first-generation detectors, which have already operated; second-generation detectors currently under construction; and third-generation detectors at the planning stage.

The most notable ground-based detectors are LIGO and Virgo, which work in collaboration, supported by GEO600. LIGO and Virgo have completed science runs as first-generation detectors, now referred to as Initial LIGO (iLIGO) and Initial Virgo (iVirgo). They are currently being upgraded to second-generation detectors, advanced LIGO (aLIGO) and advanced Virgo (AdV). LIGO has two observatories: one at Hanford, Washington, which has two detectors; and another at Livingston, Louisiana. There is an agreement to move one of the upgraded Hanford detector systems to a location in India \citep{LIGO-India,Unnikrishnan2013}. The advanced detectors should start operation in the next couple of years, with LIGO-India following further in the future.

TAMA300 is a Japanese first-generation detector. Its successor is the Kamioka Gravitational Wave Detector (KAGRA), formerly the Large-scale Cryogenic Gravitational wave Telescope (LCGT), which is located underground in the Kamioka mine. It employs more sophisticated noise-reduction techniques than LIGO or Virgo, such as cryogenic cooling.

The Einstein Telescope (ET) is an ambitious proposal to construct an underground third-generation detector. Its location would provide shielding from seismic noise, allowing it to observe frequencies of $10$--$10^4~\mathrm{Hz}$.

@@

We use an interpolation to the data published on \url{https://wwwcascina.virgo.infn.it/advirgo/} (2013) for the AdV sensitivity curve, an interpolation to the data for version D of the KAGRA detector published on \url{http://gwcenter.icrr.u-tokyo.ac.jp/en/researcher/parameter} (2013) and analytic fits to the sensitivity curves from \citet{Sathyaprakash} for the remaining detectors.

@@

These missions are designed to probe the decihertz region of the GW spectrum; they are considerably more ambitious than the LISA or eLISA mission and their launches will be further into the future. We use a simple analytic fit to the sensitivity curve for ALIA~\citep{bender_possible_2013}, while for DECIGO and BBO, fits to the sensitivity curves given by \citet{2011PhRvD..83d4011Y} are used.

### 2014-07-08 Introduction.tex
Odds and Ends

For simplicity, it is assumed that the noise in the GW detector is stationary and Gaussian (with zero mean); under these assumptions the noise is fully characterised via the one-sided noise power spectral density (PSD) $S_{n}(f)$, 
\begin{equation}\label{eq:psd}
\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>=\frac{1}{2}\delta (f-f')S_{n}(f) \; ,
\end{equation}
where angle brackets $\left<\ldots\right>$ denote an ensemble average over many noise realisations \citep{Cutler1994}. In reality, we have only a realisation to work with, however the ensemble average can replaced by a time average for stationary stochastic noise. The procedure is to measure the noise over a sufficiently long duration $T$ and then compute the Fourier transform $\tilde{n}(f)$ with a frequency resolution $\Delta f = T^{-1}$; this is repeated many times to give an average. The noise PSD $S_{n}(f)$ has units of inverse frequency.

@@

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting by taking the time average of the square of the detector noise:
\begin{eqnarray}\label{eq:meansquare}
\overline{\left|n(t)\right|^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right) \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau) \right\}(f')\right]^{*} \nonumber \\*
 & & \times \exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right)
\end{eqnarray}
Where we have substituted in using the definitions of the Fourier transform and its inverse. A property of Fourier transforms is that in the a time-domain translation by amount $t$ translates to a frequency-domain phase change $2\pi ft$; if ${\mathcal{F}}\left\{ n(\tau) \right\}(f) = \tilde{n}(f)$, then ${\mathcal{F}}\left\{ n(\tau-t) \right\}(f) = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore, the exponential factors in (\ref{eq:meansquare}) may be absorbed as
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &=& \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T} \mathrm{d}t\:\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau-t) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau-t) \right\}(f')\right]^{*} \, .
\end{eqnarray}
Since the noise is a randomly varying signal, we can use the ergodic principle to equate a time average, denoted by $\overline{(\ldots)}$, with an ensemble average, denoted by $\left<\ldots\right>$. The noise is stationary, consequently, its expectation value is unchanged by time-translation performed above. Therefore,
\begin{eqnarray}\label{eq:meansquare1}
\overline{\left|n(t)\right|^{2}} &=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right> \\
&=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
&=& \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f),
\end{eqnarray}
using (\ref{eq:psd}).

>>>>

For simplicity, it is assumed that the noise in the GW detector is stationary and Gaussian (with zero mean); under these assumptions the noise is fully characterised via the one-sided noise power spectral density (PSD) $S_{n}(f)$, 
\begin{equation}\label{eq:psd}
\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>=\frac{1}{2}\delta (f-f')S_{n}(f) \; ,
\end{equation}
where angle brackets $\left<\ldots\right>$ denote an ensemble average over many noise realisations \citep{Cutler1994}. In reality, we have only a single realisation to work with, however the ensemble average can replaced by a time average for stationary stochastic noise. The procedure is to measure the noise over a sufficiently long duration $T$ and then compute the Fourier transform $\tilde{n}(f)$ with a frequency resolution $\Delta f = T^{-1}$; this is repeated many times to give an average. The noise PSD $S_{n}(f)$ has units of inverse frequency.

@@

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting by taking the time average of the square of the detector noise:
\begin{eqnarray}\label{eq:meansquare}
\overline{\left|n(t)\right|^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right) \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau) \right\}(f')\right]^{*} \nonumber \\*
 & & \times \exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right),
\end{eqnarray}
where we have substituted in using the definitions of the Fourier transform and its inverse. A property of Fourier transforms is that a time-domain translation by amount $t$ is equivalent to a frequency-domain phase change $2\pi ft$; if ${\mathcal{F}}\left\{ n(\tau) \right\}(f) = \tilde{n}(f)$, then ${\mathcal{F}}\left\{ n(\tau-t) \right\}(f) = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore, the exponential factors in (\ref{eq:meansquare}) may be absorbed as
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &=& \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T} \mathrm{d}t\:\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau-t) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau-t) \right\}(f')\right]^{*} \, .
\end{eqnarray}
Since the noise is a randomly varying signal, we can use the ergodic principle to equate a time average, denoted by $\overline{(\ldots)}$, with an ensemble average, denoted by $\left<\ldots\right>$. The noise is stationary, consequently, its expectation value is unchanged by time-translation performed above. Therefore,
\begin{eqnarray}\label{eq:meansquare1}
\overline{\left|n(t)\right|^{2}} &=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right> \\
&=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
&=& \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f),
\end{eqnarray}
using (\ref{eq:psd}).

### 2014-07-08 Plots.tex
Ending
Odds and Ends

\appendix
\section{Sensitivity curves}\label{app:a}
The plots in this section show all of the detectors and sources described in the text. Interactive versions of these plots may be viewed on-line, \url{http://www.ast.cam.ac.uk/~rhc26/sources/}. The detector noise curves all have their resonance spikes removed for clarity. 
\begin{figure}[h!]
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure1.pdf}
 \caption{A plot of characteristic strain against frequency for a variety of detectors and sources.}
 \label{fig:hc}
\end{figure}
\begin{figure}[h!]
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure2.pdf}
 \caption{A plot of the square root of power spectral density against frequency for a variety of detectors and sources.}
 \label{fig:S}
\end{figure}
\begin{figure}[h!]
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure3.pdf}
 \caption{A plot of the dimensionless energy density in GWs against frequency for a variety of detectors and sources.}
 \label{fig:omega}
\end{figure}

>>>>

\appendix
\section{Sensitivity curves}\label{app:a}
The plots in this section show all of the detectors and sources described in the main text. Interactive versions of these plots, allowing for removal of any of the curves, may be created and downloaded on-line, \url{http://www.ast.cam.ac.uk/~rhc26/sources/}. The detector noise curves all have their resonance spikes removed for clarity. 
\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure1.pdf}
 \caption{A plot of characteristic strain against frequency for a variety of detectors and sources.}
 \label{fig:hc}
\end{figure}
\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure2.pdf}
 \caption{A plot of the square root of power spectral density against frequency for a variety of detectors and sources.}
 \label{fig:S}
\end{figure}
\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.85\textwidth]{figure3.pdf}
 \caption{A plot of the dimensionless energy density in GWs against frequency for a variety of detectors and sources.}
 \label{fig:omega}
\end{figure}

### 2014-07-08 Sources.tex
Odds and Ends

Simulations of core-collapse supernovae show that GWs between $(10^{2}$--$10^{3})~\mathrm{Hz}$ can be produced \citep{Kotake2006}. The GW signal undergoes ${\mathcal{O}}(1)$ oscillation and is hence burst-like. \citet{2002A&A...393..523D} calculate the average maximum amplitude of GWs for a supernova at distance $r$ as
\begin{equation}
h=8.9\times 10^{-21}\left( \frac{10~\mathrm{kpc}}{r} \right) \; .
\end{equation}
The event rate for supernovae is approximately $\rho_{\mathrm{SN}} = 5\times10^{-4}~\mathrm{Mpc^{-3}\,yr^{-1}}$. The boxes plotted in \ref{app:a} correspond to a distance $r=3~\mathrm{Mpc}$ with the frequency range quoted above. The LIGO and Virgo detectors have already placed bounds on the event rate for these sources \citep{Bursts}.

@@

Rotating neutron stars are a source of continuous GWs if they possess some degree of axial asymmetry \citep{Abbott2007, Prix2009, Einstein@Home}. The signals are near monochromatic with a frequency twice the rotation frequency of the neutron star, and are a potential source for ground-based detectors. The amplitude of the GWs depends upon the deformation of the neutron star, which itself depends upon the neutron star equation of state. This is currently uncertain \citep{Lattimer2012}. Several known pulsars could be sources for the advanced detectors and upper limits from the initial detectors help to constrain the deformations \citep{Aasi2014a}. The boxes plotted in \ref{app:a} is the same as for supernovae \cplb{for reasons I do not entirely understand}.

>>>>

Simulations of core-collapse supernovae show that GWs between $(10^{2}$--$10^{3})~\mathrm{Hz}$ can be produced \citep{Kotake2006}. The GW signal undergoes ${\mathcal{O}}(1)$ oscillation and is hence burst-like. \citet{2002A&A...393..523D} calculate the average maximum amplitude of GWs for a supernova at distance $r$ as
\begin{equation}
h=8.9\times 10^{-21}\left( \frac{10~\mathrm{kpc}}{r} \right) \; .
\end{equation}
The event rate for supernovae is approximately $\rho_{\mathrm{SN}} = 5\times10^{-4}~\mathrm{Mpc^{-3}\,yr^{-1}}$. The boxes plotted in \ref{app:a} correspond to a distance $r=300~\mathrm{kpc}$ with the frequency range quoted above. The LIGO and Virgo detectors have already placed bounds on the event rate for these sources \citep{Bursts}.

@@

Rotating neutron stars are a source of continuous GWs if they possess some degree of axial asymmetry \citep{Abbott2007, Prix2009, Einstein@Home}. The signals are near monochromatic with a frequency twice the rotation frequency of the neutron star, and are a potential source for ground-based detectors. The amplitude of the GWs depends upon the deformation of the neutron star, which itself depends upon the neutron star equation of state. This is currently uncertain \citep{Lattimer2012}. Several known pulsars could be sources for the advanced detectors and upper limits from the initial detectors help to constrain the deformations. The boxes plotted in \ref{app:a} correspond to the upper limits placed on a GW signal from the Crab pulsar \citep{Aasi2014a}, extrapolated across a frequency range between $(20$--$10^{3})~\mathrm{Hz}$.
